---
title: "Lecture 4a"
subtitle: "Basic plotting in R"
output: slidy_presentation
date: "2023-10-27"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

## Basic plotting with R

- Basic plotting
- Title, labels, legend, coordinate system
- Histogram
- Output devices
- ggplot2

## Basic plotting

Prepare some data for plotting

```{r}
my.data <- read.csv("https://raw.githubusercontent.com/mikdur/r-primer/main/lecture1/example.csv")
my.data$mass <- my.data$len ^ 2 * 8 + rnorm( dim(my.data)[1], mean=1, sd=10)
my.data$mature <- my.data$age > 4
summary(my.data)
```

## Basic plotting

Plot everything against everything

```{r}
plot(my.data)
```

## Simple X/Y scatter plot

```{r}
plot( my.data$age, my.data$mass )
```

## Simple X/Y scatter plot

```{r}
plot( my.data$age, my.data$mass,
      xlab = "Age",
      ylab = "Mass")
```

## Simple X/Y scatter plot

```{r}
plot( my.data$age, my.data$mass,
      xlab = "Age",
      ylab = "Mass",
      main = "Mass developen of the Splonk")
```

## Simple X/Y scatter plot

```{r}
plot( my.data$age, my.data$mass,
      xlab = "Age",
      ylab = "Mass",
      main = "Mass developen of the Splonk",
      xlim = c(1, 10) )
```

## Simple X/Y scatter plot

```{r}
plot( my.data$age, my.data$mass,
      xlab = "Age",
      ylab = "Mass",
      main = "Mass developen of the Splonk",
      xlim = c(1, 10))
lines(my.data$age, my.data$mass)
```


## Boxplot

```{r}
par(mfcol=c(1,2))
plot(mass ~ mature, data=my.data)
plot(mass ~ as.factor(mature), data=my.data)
```


## Coordinate system

- Coodrinate system with origin in bottom left corner
- Scaled to fit plotted values (unless overridden)

```{r}
x <- c(0, 0, 0, 1, 1, 1, 2, 2, 3, 2, 2, 3, 2, 2, 4, 4, 4, 5, 5)
y <- c(0, 2, 1, 1, 2, 0, 2, 2, 2, 2, 1, 2, 1, 0, 0, 2, 0, 0, 1)

plot(x, y, pch=1)
```

## Coordinate system

- Coodrinate system with origin in bottom left corner
- Scaled to fit plotted values (unless overridden)

```{r}
x <- c(0, 0, 0, 1, 1, 1, 2, 2, 3, 2, 2, 3, 2, 2, 3, 4, 4, 4, 5, 5)
y <- c(0, 2, 1, 1, 2, 0, 0, 2, 2, 2, 1, 1, 1, 0, 0, 0, 2, 0, 0, 1)

plot(x, y)
lines(x, y)
```

## Coordinate system

- Coodrinate system with origin in bottom left corner
- Scaled to fit plotted values (unless overridden)

```{r}
x <- c(0, 0, 0, 1, 1, 1, 2, 2, 3, 2, 2, 3, 2, 2, 3, 4, 4, 4, 5, 5)
y <- c(0, 2, 1, 1, 2, 0, 0, 2, 2, 2, 1, 1, 1, 0, 0, 0, 2, 0, 0, 1)

plot(x, y)
lines(x, y)
text( x + 0.2, y, labels = as.character( 1:20 ) )
legend("topright", pch=1:3, legend = c("first", "second", "third"))
```

## Plotting characters

- pch argument defines the plotting character

```{r}
plot(1:10, 1:10, pch = 1:10)
legend("bottomright", pch = 1:10, legend = as.character(1:10))
```

## Histograms

```{r}
x <- rnorm(1000)
hist(x)
```

## Histograms

```{r}
x <- rnorm(1000)
hist(x, breaks = 40)
```

## Histograms

```{r}
x <- rnorm(1000)
hist(x, breaks = 40, freq=FALSE)
x.density <- density(x)
lines(x.density)
```

## Output devices

R can create high quality vector graphics various formats
- PDF (pdf())
- EPS (postscript())
- SVG (svg())

To create a pdf/eps you first need to open an output device, then plot to it:

```{r}
pdf("foo.pdf", width = 4, height = 4)
plot(1:10, 20:11)
dev.off()
```

## ggplot2 - tidyverse plotting

ggplot2 provides a different take on plotting. Rather than specifying each and 
every detail yourself, it provides means of telling what to plot illustrate, and 
then takes care of the details.

Basically, you provide what you want to plot:

- Data ((tidy) dataframe)
- Aestethics mapping
- How to plot (geometry)

## ggplot2 - Example

Let's try with our toy dataset:

```{r}
ggplot( data=my.data, 
        aes(age, len, colour = as.factor(mature))) +
  geom_point()
```

## Faceting

```{r}
ggplot( data=my.data, 
        aes(age, len, colour = as.factor(mature))) +
  geom_point() +
  facet_wrap(~ as.factor(mature))
```

## Multiple panels

```{r}
library(patchwork)

plot.1 <- ggplot( data=filter(my.data, mature == 1), 
                  aes(age, len, colour = as.factor(mature))) +
            geom_point()
plot.2 <- ggplot( data=filter(my.data, mature == 0), 
                  aes(age, len, colour = as.factor(mature))) +
            geom_point()
plot.3 <- ggplot( data=filter(my.data, mature == 1), 
                  aes(age, mass, colour = as.factor(mature))) +
            geom_point()
plot.4 <- ggplot( data=filter(my.data, mature == 0), 
                  aes(age, mass, colour = as.factor(mature))) +
            geom_point()

(plot.1 + plot.2) / (plot.3 + plot.4)

```

## Further reading

- Help
- R Graphics Cookbook https://r-graphics.org
- Visualisation chapters in "R for Data Science" https://r4ds.had.co.nz


